Trending assessment for nowcasts, measurements, and forecasts

### 2024-03-03 sections/03-trending.tex
Update on Overleaf.

Second, in the context of nowcasting or forecasting, several signals for the same time $t + l$ are often available at different times of generation.
Thus, it has to be specified in notation which signal generated at which time is used.
We use the term \textit{issue time} for the generation time of a signal and \textit{target time} to denote the time the forecast applies to.

For measurement, usually, a test method is evaluated against a gold standard to test whether the test method could replace the gold standard.
If the test method is assessed to be consistent with the gold standard, the gold standard measurement values are not available to compute the predicted change.
Thus, the predicted change is computed based on the test method's values at time $t$ and $t + l$.

@@

In addition to these straight shapes, exclusion areas such as ellipsis could have more difficult shapes.
Such a shape could, for example, be obtained by using a multivariate normal error model around the zero point and excluding points, that exceed a certain likelihood of being obtained by chance from the zero point.
These shapes are theoretically appealing, but the interpretation of the resulting estimators gets complex.
Thus, we use the simple exclusion areas above.

In most applications, it is advisable to determine the exact shape and size of the exclusion area based on domain knowledge or expert opinions.
In addition, the size can also be calculated as a proportion of the total variance or the total range of the data.
A third approach is to visualize the trending ratio for different values of $e$.
In Section~\ref{sec:application}, there are various examples of such a plot.
As a base approach, we advocate using no exclusion area in the four-quadrant plot as the points do not complicate its interpretation and a $\acceps$-over-$e$ plot to visualize the course of the trending ratio over different exclusion area sizes.

\subsection{The conditional trending plot and confidence intervals based on bootstrapping}\label{subsec:trending-cond-prob}
The above-defined estimators give information on the probabilites $P(\diffxrv \diffyrv > 0 | \diffxrv \diffyrv \notin e)$, $P(\diffxrv \diffyrv > 0 | \diffxrv > 0, \diffxrv \diffyrv \notin e)$ and $P(\diffxrv \diffyrv > 0 | \diffxrv < 0, \diffxrv \diffyrv \notin e)$.
These probabilites point a general picture, but still too coarse for local effects.
To assess the local trending ability of a prediction, more insights might be gained by considering the conditional distribution.
Whereas it would be possible to build analog measures on other intervals, assessing $P(\diffxrv \diffyrv > 0 | \diffx = x)$ graphically eases the simultaneous evaluation of various intervals.
In addition, the conditional trending plot facilitates the comparison of various methods in a single plot, and asymmetries of $P(\diffxrv \diffyrv > 0 | \diffx = x)$ with respect to $x$ in the trending ability can be detected.

A multivariate \acf{kde} facilitates a continuous estimation of $P(\diffxrv \diffyrv > 0 | \diffxrv = x)$ by estimating the components $f_{\diffxrv, \diffyrv}$ and $f_{\diffxrv}$ of
\begin{align*}
P(\diffxrv \diffyrv > 0 | \diffxrv = x) = \begin{cases}
                                              \int_{-\infty}^0 \frac{f_{\diffxrv, \diffyrv}(x, y)}{f_{\diffxrv}(x)} \: \textrm{d} \: y & \text{if } x < 0, \\
                                              \int_{0}^{\infty} \frac{f_{\diffxrv, \diffyrv}(x, y)}{f_{\diffxrv}(x)} \: \textrm{d} \: y & \text{if } x > 0, \\
\end{cases}
\end{align*}
for $x \neq 0$ through a \ac{kde}.
A comprehensive introduction into multivariate \ac{kde} can be found in \textcite{Gramacki2018}, and implementations are available in many programming languages~\parencite[e.g., for  Python in][]{Seabold2010}.
The \ac{kde} yields estimates for $P(\diffxrv \diffyrv > 0 | \diffxrv = x)$ for all values of $x \in \R$.
Note that for kernels with infinite support, for example, the Gaussian kernel, the plot should be limited to the core values of $\diffx$ without outliers to avoid estimates based on small subsets of the data.
Besides the kernel, a bandwidth selector has to be chosen for a multivariate \ac{kde}.
Various methods are available with different strengths and weaknesses.
Figure~\ref{fig:trending-cond-prob-bw} shows the resulting conditional trending plots for the three well-known selectors, rule-of-thumb, cross-validation maximum likelihood, and cross-validation least squares using the \verb|statsmodels| python package~\parencite{Seabold2010}.
While the rule-of-thumb is based only on the covariance matrix, the other two numerically optimize the bandwidth with a hold-one-out least squares or likelihood objective function.
The dashed line shows the theoretical $P(\diffyrv \diffxrv > 0 | \diffxrv = x)$.
The second method, cross-validation least squares, needs long computation times while yielding small or no bandwidth results, even for the two relatively small datasets.
The rule-of-thumb and cross-validation maximum likelihood both yield reasonable results at moderate computation times. 
Further examples, including comparisons between methods concerning their trending ability, are available for the applications in Section~\ref{sec:application}.

\begin{figure}
    \centering
    \begin{subfigure}{.48\textwidth}
        \includegraphics{plots/illustrative_examples/cond_prob_plot_bw_asym_butterfly}
        \caption{First dataset with asymmetric dependence.}
    \end{subfigure}
    \begin{subfigure}{.48\textwidth}
        \includegraphics{plots/illustrative_examples/cond_prob_plot_bw_normal}
        \caption{Second dataset. }
    \end{subfigure}
    \caption{Resulting conditional trending plot for different bandwidth selection processes. Cross-validation least squares takes a considerably larger computation time. It does not converge for the first data set and yields a too small bandwidth for the second data set. The rule of thumb is the fastest method but tends to oversmooth. The cross-validation maximum likelihood method yields a more reasonable bandwidth with moderate computation time. }\label{fig:trending-cond-prob-bw}
\end{figure}

\todo{Übergang zu Bootstrapping}
Confidence intervals can be used to account for the estimation uncertainty of the measures above.
Bootstrap confidence intervals are a nonparametric technique based on resampling~\parencite[for introductions see][]{Hesterberg2011,Bittmann2021}.
New samples are drawn with replacement from the dataset in bootstrapping and are not based on parametric assumptions as for classical confidence intervals.
The confidence interval is then computed based on the bootstrap samples.
We examine three methods for bootstrapping here: the intuitive percentile and the more sophisticated basic and \ac{bca} method.
In the \textit{percentile} approach, the confidence interval for level $\alpha$ is built directly from the bootstrap estimators' empirical distribution.
The \textit{basic} approach computes the confidence interval based on the non-bootstrapped estimate using the bootstrapped quantile deviations~\parencite{Davison1997}.
The \ac{bca} method modifies the quantiles of the empirical bootstrap distribution by a bias and an acceleration parameter~\parencite{Efron1987}.
Typically, the percentile approach needs larger datasets and provides an easy and fast estimate, while the \ac{bca} is computationally expensive but needs smaller datasets for reasonable confidence intervals.
The basic approach balances those two aims.
We compare the approaches in a small synthetic data study concerning their small-dataset behavior and computation time in Appendix~\ref{sec:appendix}.
For small datasets, the \ac{bca} holds the confidence level also for small data sets with a neglibly higher computation time than the other approaches.
Thus, we use the \ac{bca} method for the confidence intervals in the applications in Section~\ref{sec:application}.

In standard bootstrapping, the sample is assumed to be independent and identically distributed.
Thus, before applying bootstrapping methods, the strength of sequential dependence should be inspected, for example, by analyzing the autocorrelation and partial autocorrelation.
The simple bootstrapping methods above do not account for serial dependence. 
Bootstrapping focusing on time series data is covered in~\textcite{Hardle2003,Kreiss2012}, for example.

>>>>

Second, in the context of nowcasting or forecasting, several signals for the same time $t + l$ are often available at different generation times.
Thus, it has to be specified in notation which signal generated at which time is used.
We use the term \textit{issue time} for the generation time of a signal and \textit{target time} to denote the time the forecast applies to.

For measurement, a test method is usually evaluated against a gold standard to test whether the test method could replace the gold standard.
If the test method is assessed to be consistent with the gold standard, the gold standard measurement values are not available to compute the predicted change.
Thus, the predicted change is computed based on the test method's values at time $t$ and $t + l$.

@@

In addition to these straight shapes, exclusion areas could have more complex shapes, such as ellipses.
Such a shape could, for example, be obtained by using a multivariate normal error model around the zero point and excluding points that exceed a certain likelihood of being produced by chance from the zero point.
These shapes are theoretically appealing, but interpreting the resulting estimators gets complex.
Thus, we use the simple exclusion areas above.

In most applications, it is advisable to determine the exact shape and size of the exclusion area based on domain knowledge or expert opinions.
In addition, the size can also be calculated as a proportion of the total variance or the total range of the data.
A third approach is to visualize the trending ratio for different values of $e$.
Section~\ref{sec:application} contains various examples of such a plot.
As a base approach, we advocate using no exclusion area in the four-quadrant plot as the points do not complicate its interpretation and a $\acceps$-over-$e$ plot to visualize the course of the trending ratio over different exclusion area sizes.

\subsection{The conditional trending plot and confidence intervals based on bootstrapping}\label{subsec:trending-cond-prob}
The above-defined estimators give information on the probabilites $P(\diffxrv \diffyrv > 0 | \diffxrv \diffyrv \notin e)$, $P(\diffxrv \diffyrv > 0 | \diffxrv > 0, \diffxrv \diffyrv \notin e)$ and $P(\diffxrv \diffyrv > 0 | \diffxrv < 0, \diffxrv \diffyrv \notin e)$.
These probabilities draw a general picture but might still be too coarse for local effects.
More insights might be gained by considering the conditional distribution to assess the local trending ability of a prediction.
Whereas it would be possible to build analog measures on other intervals, assessing $P(\diffxrv \diffyrv > 0 | \diffx = x)$ graphically eases the simultaneous evaluation of various intervals.
In addition, the conditional trending plot facilitates the comparison of various methods in a single plot and asymmetries of $P(\diffxrv \diffyrv > 0 | \diffx = x)$ with respect to $x$ in the trending ability can be detected.

A multivariate \acf{kde} facilitates a continuous estimation of $P(\diffxrv \diffyrv > 0 | \diffxrv = x)$ by estimating the components $f_{\diffxrv, \diffyrv}$ and $f_{\diffxrv}$ of
\begin{align*}
P(\diffxrv \diffyrv > 0 | \diffxrv = x) = \begin{cases}
                                              \int_{-\infty}^0 \frac{f_{\diffxrv, \diffyrv}(x, y)}{f_{\diffxrv}(x)} \: \textrm{d} \: y & \text{if } x < 0, \\
                                              \int_{0}^{\infty} \frac{f_{\diffxrv, \diffyrv}(x, y)}{f_{\diffxrv}(x)} \: \textrm{d} \: y & \text{if } x > 0, \\
\end{cases}
\end{align*}
for $x \neq 0$ through a \ac{kde}.
A comprehensive introduction into multivariate \ac{kde} can be found in \textcite{Gramacki2018}, and implementations are available in many programming languages~\parencite[e.g., for  Python in][]{Seabold2010}.
The \ac{kde} yields estimates for $P(\diffxrv \diffyrv > 0 | \diffxrv = x)$ for all values of $x \in \R$.
Note that for kernels with infinite support, for example, the Gaussian kernel, the plot should be limited to the core values of $\diffx$ without outliers to avoid estimates based on small subsets of the data.
Besides the kernel, a bandwidth selector has to be chosen for a multivariate \ac{kde}.
Various methods are available with different strengths and weaknesses.
Figure~\ref{fig:trending-cond-prob-bw} shows the resulting conditional trending plots for the three well-known selectors, rule-of-thumb, cross-validation maximum likelihood, and cross-validation least squares using the \verb|statsmodels| python package~\parencite{Seabold2010}.
While the rule-of-thumb is based only on the covariance matrix, the other two numerically optimize the bandwidth with a hold-one-out least squares or likelihood objective function.
The dashed line shows the theoretical $P(\diffyrv \diffxrv > 0 | \diffxrv = x)$.
The second method, cross-validation least squares, needs long computation times while yielding small or no bandwidth results, even for two relatively small datasets.
The rule-of-thumb and cross-validation maximum likelihood both yield reasonable results at moderate computation times. 
Further examples, including comparisons between methods concerning their trending ability, are available for the applications in Section~\ref{sec:application}.

\begin{figure}
    \centering
    \begin{subfigure}{.48\textwidth}
        \includegraphics{plots/illustrative_examples/cond_prob_plot_bw_asym_butterfly}
        \caption{First dataset with asymmetric dependence.}
    \end{subfigure}
    \begin{subfigure}{.48\textwidth}
        \includegraphics{plots/illustrative_examples/cond_prob_plot_bw_normal}
        \caption{Second dataset. }
    \end{subfigure}
    \caption{Resulting conditional trending plot for different bandwidth selection processes. Cross-validation least squares takes a considerably larger computation time. It does not converge for the first data set and yields a bandwidth too small for the second data set. The rule of thumb is the fastest method but tends to oversmooth. The cross-validation maximum likelihood method yields a more reasonable bandwidth with moderate computation time. }\label{fig:trending-cond-prob-bw}
\end{figure}

\todo{Übergang zu Bootstrapping}
Confidence intervals can account for the estimation uncertainty of the measures above.
Bootstrap confidence intervals are a nonparametric technique based on resampling~\parencite[for introductions see][]{Hesterberg2011,Bittmann2021}.
New samples are drawn with replacement from the dataset in bootstrapping and are not based on parametric assumptions as for classical confidence intervals.
The confidence interval is then computed based on the bootstrap samples.
We examine three methods for bootstrapping here: the intuitive percentile and the more sophisticated basic and \ac{bca} method.
In the \textit{percentile} approach, the confidence interval for level $\alpha$ is built directly from the bootstrap estimators' empirical distribution.
The \textit{basic} approach computes the confidence interval based on the non-bootstrapped estimate using the bootstrapped quantile deviations~\parencite{Davison1997}.
The \ac{bca} method modifies the quantiles of the empirical bootstrap distribution by a bias and an acceleration parameter~\parencite{Efron1987}.
Typically, the percentile approach needs larger datasets and provides an easy and fast estimate, while the \ac{bca} is computationally expensive but needs smaller datasets for reasonable confidence intervals.
The basic approach balances those two aims.
We compare the approaches in a small synthetic data study concerning their small-dataset behavior and computation time in Appendix~\ref{sec:appendix}.
For small datasets, the \ac{bca} holds the confidence level also for small data sets with a negligibly higher computation time than the other approaches.
Thus, we use the \ac{bca} method for the confidence intervals in the applications in Section~\ref{sec:application}.

The sample is assumed to be independent and identically distributed in standard bootstrapping.
Thus, before applying bootstrapping methods, the strength of sequential dependence should be inspected, for example, by analyzing the autocorrelation and partial autocorrelation.
The simple bootstrapping methods above do not account for serial dependence. 
Bootstrapping focusing on time series data is covered in~\textcite{Hardle2003,Kreiss2012}, for example.
